---
title: "Introduction to the risks package: Get Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the risks package: Get Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Installation

The **risks** package can be installed from CRAN:

```{r cran_install, eval = FALSE}
install.packages("risks")
```

Development versions can be installed from 
[GitHub](https://stopsack.github.io/risks/):

```{r github_install, eval = FALSE}
remotes::install_github("stopsack/risks")
```

# An example cohort study

We use a cohort of women with breast cancer, as used by Spiegelman and Hertzmark ([Am J Epidemiol 2005](https://pubmed.ncbi.nlm.nih.gov/15987728)) and Greenland ([Am J Epidemiol 2004](https://pubmed.ncbi.nlm.nih.gov/15286014)). The the categorical exposure is `stage`, the binary outcome is `death`, and the binary confounder is `receptor`.

```{r load, message = FALSE}
library(risks)  # provides riskratio(), riskdiff(), postestimation functions
library(dplyr)  # For data handling
library(broom)  # For tidy() model summaries

data(breastcancer)  # Load sample data

breastcancer %>%  # Display the sample data
  group_by(receptor, stage) %>% 
  summarize(
    deaths = sum(death), 
    total = n(), 
    risk = deaths / total
  )
```


# Adjusted risk ratios

The risk of death is higher among women with higher-stage and hormone receptor-low cancers, which also tend to be of higher stage. Using `risks` models to obtain (possibly multivariable-adjusted) risk ratios or risk differences is similar to the standard code for logistic models in R. As customary in R, log(RR) is returned; see below for how to obtain exponentiated values. No options for model `family` or `link` need to be supplied:

```{r basic}
fit_rr <- riskratio(
  formula = death ~ stage + receptor, 
  data = breastcancer
)
summary(fit_rr)
```

  
# Adjusted risk differences

To obtain risk differences, use `riskdiff`, which has the same syntax:

```{r basic2}
fit_rd <- riskdiff(
  formula = death ~ stage + receptor, 
  data = breastcancer
)
summary(fit_rd)
```

For example, the risk of death was 57 percentage points higher in women with stage III breast cancer compared to stage I (95% confidence interval, 38 to 77 percentage points), adjusting for hormone receptor status.

The model summary in `riskratio()` and `riskdiff()` includes to two additions compared to a regular `glm()` model:

*  In the first line of `summary()`, the type of risk regression model is displayed (in the example, "`Risk ratio model, fitted as binomial model...`"). 
*  At the end of the output, confidence intervals for the model coefficients are printed.


# Accessing model coefficients

The risks package provides an interface to `broom::tidy()`, which returns a data frame of all coefficients (risk differences in this example), their standard errors, *z* statistics, and confidence intervals.

```{r basic3}
tidy(fit_rd)
```

\
In accordance with `glm()` standards, coefficients for relative risks are shown on the logarithmic scale. Exponentiated coefficients for risk ratios are easily obtained via `tidy(..., exponentiate = TRUE)`:

```{r basic4}
tidy(
  fit_rr, 
  exponentiate = TRUE
)
```

For example, the risk of death was 6 times higher in women with stage III breast cancer compared to stage I (95% confidence interval, 3 to 13 times), adjusting for hormone receptor status.


# More post-estimation commands

Typical R functions that build on regression models can further process fitted `risks` models. In addition to `tidy()`, examples are:

*  `coef(fit)` or `coefficients(fit)` return model coefficients (*i.e.*, RDs or log(RR)) as a numeric vector
*  `confint(fit, level = 0.9)` returns *90%* confidence intervals.
*  `predict(fit, type = "response")` or `fitted.values(fit)` return predicted probabilities of the binary outcome.
*  `residuals(fit)` returns model residuals.
